Diagnostic Plots

### 2024-08-16 pre_print/pre_print_v2.qmd
Final revised v2 preprint

---
articletitle: |
  Cumulative evidence synthesis and consideration of "research waste" using Bayesian methods: An example updating a previous meta-analysis of self-talk interventions for sport/motor performance
format: 
  sportrxiv-pdf:
    include-in-header:
      text: |
       \usepackage[font=scriptsize]{caption}
       \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
author:
  - name: Hannah Corcoran
    affiliations:
      - ref: 1
  - name: James Steele
    affiliations:
      - ref: 1
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james.steele@solent.ac.uk
affiliations:
      - id: 1
        name: Department of Sport and Health, Solent University, UK
abstract: |
  In the present paper we demonstrate the application of methods for cumulative evidence synthesis including Bayesian meta-analysis, and exploration of questionable research practices such as publication bias or *p*-hacking, in the sport and exercise sciences for the evaluation of experimental interventions. The use of such methods can aid in study planning and avoid "research waste". In demonstrating and discussing these methods we use the example of self-talk interventions and their effects upon sport/motor performance given a quantitative evidence synthesis has not been conducted on this topic, to the best of our knowledge, since 2011 when Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] conducted their systematic review and meta-analysis. As such, this topic is ripe to use in demonstrating cumulative methods such as Bayesian updating. Therefore, our aim was to conduct an updated systematic review and Bayesian meta-analysis replicating the search, inclusion, and models of Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] and demonstrate the application of cumulative evidence synthesis methods including; consideration of the initial probability that a new study of the effects of self-talk interventions would shift our prior belief in their effectiveness, the application of priors taken from the previous meta-analysis to be updated by new studies identified to a new posterior estimate of effect, and consideration of other possible sources of research waste from questionable research practices such as publication bias and *p*-hacking. Such methods as those demonstrated here, when used prospectively, can aid researchers in determining whether further research of a particular experimental intervention is in fact warranted. Considering the limited resources and time for conducting research we hope that highlighting the application of these methods might help researchers in the field to avoid research waste and more productively direct their research efforts.
license-type: ccby # change if neccessary
# year: 2025 # defaults to current year
keywords: [psychological intervention, cumulative evidence synthesis, research waste] # optional
# optional link to Supplementary Material:
suppl-link: https://osf.io/dqwh5/ 
reference-section-title: References
printnote: "PREPRINT - NOT PEER REVIEWED" # if post print, include "POSTPRINT" then link to the published article
bibliography: 
  - references.bib
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

Although the literature on self-talk *interventions* and their effects on sports/motor performance specifically could be considered wasteful, as explained, the post-2011 years have seen theoretical advancements in conceptualisation and operationalisation of the construct of self-talk [@brinthauptSelftalkResearchChallenges2023; @geurtsMakingSenseSelf2018; @hardyReflectionsMaturingResearch2018; @latinjakSelfTalkInterdisciplinaryReview2023; @vanraalteSelftalkReviewSportspecific2016; @latinjakSpeakingClearly102019; @vanraalteCommonLanguageCategorization2019]. Alongside the discussion within the broader psychological literature regarding replication concerns there have also been calls to improve theorising [@eronenTheoryCrisisPsychology2021] and measurement practices [@flakeMeasurementSchmeasurementQuestionable2020]. These are important steps that researchers should consider before they move to testing hypotheses such as whether intervening on a psychological construct produces particular causal outcomes [@scheelWhyHypothesisTesters2021]. As such, it would be unfair to suggest that work aimed in these specific theoretical/operationalisation directions has been wasteful and attempts at cumulative evidence synthesis which draw such conclusions might also consider qualitative appraisal of the broader literature saving conlusions about "research waste" for the specific context explored.

@@

This work has demonstrated the application of cumulative evidence synthesis methods including; consideration of the initial probability that a new study of the effects of self-talk interventions would shift our prior belief in their effectiveness, the application of priors taken from the previous meta-analysis to be updated by new studies identified to a new posterior estimate of effect, and consideration of other sources of research waste from questionable research practices such as possible publication bias and *p*-hacking. We presented the application of these methods in the example of self-talk interventions for sports/motor performance given it has been over a decade since Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] published their meta-analysis of self-talk interventions. The application of these methods highlight that much of the literature on this topic could be considered "research waste"; the impact of any new study upon the prior estimate from Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] would have had added little evidence to change prior beliefs, the doubling of research conducted since 2011 also added little evidence to update the prior estimate, and it is likely that much of the research post-2011 (and likely prior to 2011) has been subject to questionable research practices making it unclear whether self-talk interventions affect sports/motor performance or not. This example can act as a warning to researchers in the sport and exercise sciences about what kinds of questions they should ask themselves prior to beginning their work. Such methods as those demonstrated here, when used prospectively, can aid researchers in determining whether further research of a particular experimental intervention is in fact warranted; and when used in retrospect may reveal where research has been a waste. Considering the limited resources and time for conducting research their application may suggest it to be more worthwhile moving onto other more pertinent research questions or programmes. In the case of self-talk specifically, future research might be best placed to continue with the intention of building strong theories regarding the conceptualisation and operationalisation of self-talk and, following this, test if there really is an intervention effect as hypothesised whilst incorporating safe-guards for questionable research pratices e.g., pre-registration/registered-reports [@chambersPresentFutureRegistered2022].

>>>>

---
articletitle: |
  Cumulative evidence synthesis and consideration of "research waste" using Bayesian methods: An example updating a previous meta-analysis of self-talk interventions for sport/motor performance
format: 
  sportrxiv-pdf:
    include-in-header:
      text: |
       \usepackage[font=scriptsize]{caption}
       \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
author:
  - name: Hannah Corcoran
    affiliations:
      - ref: 1
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
affiliations:
      - id: 1
        name: Department of Sport and Health, Solent University, UK
      - id: 2
        name: Steele Research Ltd, UK
abstract: |
  In the present paper we demonstrate the application of methods for cumulative evidence synthesis including Bayesian meta-analysis, and exploration of questionable research practices such as publication bias or *p*-hacking, in the sport and exercise sciences for the evaluation of experimental interventions. The use of such methods can aid in study planning and avoid "research waste". In demonstrating and discussing these methods we use the example of self-talk interventions and their effects upon sport/motor performance given a quantitative evidence synthesis has not been conducted on this topic, to the best of our knowledge, since 2011 when Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] conducted their systematic review and meta-analysis. As such, this topic is ripe to use in demonstrating cumulative methods such as Bayesian updating. Therefore, our aim was to conduct an updated systematic review and Bayesian meta-analysis replicating the search, inclusion, and models of Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] and demonstrate the application of cumulative evidence synthesis methods including; consideration of the initial probability that a new study of the effects of self-talk interventions would shift our prior belief in their effectiveness, the application of priors taken from the previous meta-analysis to be updated by new studies identified to a new posterior estimate of effect, and consideration of other possible sources of research waste from questionable research practices such as publication bias and *p*-hacking. Such methods as those demonstrated here, when used prospectively, can aid researchers in determining whether further research of a particular experimental intervention is in fact warranted. Considering the limited resources and time for conducting research we hope that highlighting the application of these methods might help researchers in the field to avoid research waste and more productively direct their research efforts.
license-type: ccby # change if neccessary
# year: 2025 # defaults to current year
keywords: [psychological intervention, cumulative evidence synthesis, research waste] # optional
# optional link to Supplementary Material:
suppl-link: https://osf.io/dqwh5/ 
reference-section-title: References
printnote: "PREPRINT - NOT PEER REVIEWED" # if post print, include "POSTPRINT" then link to the published article
bibliography: 
  - references.bib
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

Although the literature on self-talk *interventions* and their effects on sports/motor performance specifically could be considered wasteful, as explained, the post-2011 years have seen theoretical advancements in conceptualisation and operationalisation of the construct of self-talk [@brinthauptSelftalkResearchChallenges2023; @geurtsMakingSenseSelf2018; @hardyReflectionsMaturingResearch2018; @latinjakSelfTalkInterdisciplinaryReview2023; @vanraalteSelftalkReviewSportspecific2016; @latinjakSpeakingClearly102019; @vanraalteCommonLanguageCategorization2019]. Alongside the discussion within the broader psychological literature regarding replication concerns there have also been calls to improve theorising [@eronenTheoryCrisisPsychology2021] and measurement practices [@flakeMeasurementSchmeasurementQuestionable2020]. These are important steps that researchers should consider before they move to testing hypotheses such as whether intervening on a psychological construct produces particular causal outcomes [@scheelWhyHypothesisTesters2021]. As such, it would be unfair to suggest that work aimed in these specific theoretical/operationalisation directions has been wasteful and attempts at cumulative evidence synthesis which draw such conclusions might also consider qualitative appraisal of the broader literature saving conclusions about "research waste" for the specific context explored. But, whilst it has been argued that cumulative evidence synthesis methods such as meta-analysis to test theoretically derived hypotheses [@haggerTestingTheoryPredictions2024], for self-talk interventions specifically it may be worth starting afresh to determine if there really is an intervention effect whilst incorporating safe-guards for such issues as the questionable research practices explored here e.g., pre-registration/registered-reports [@chambersPresentFutureRegistered2022].

@@

This work has demonstrated the application of cumulative evidence synthesis methods including; consideration of the initial probability that a new study of the effects of self-talk interventions would shift our prior belief in their effectiveness, the application of priors taken from the previous meta-analysis to be updated by new studies identified to a new posterior estimate of effect, and consideration of other sources of research waste from questionable research practices such as possible publication bias and *p*-hacking. We presented the application of these methods in the example of self-talk interventions for sports/motor performance given it has been over a decade since Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] published their meta-analysis of self-talk interventions. The application of these methods highlight that much of the literature on this topic could be considered "research waste"; the impact of any new study upon the prior estimate from Hatzigeorgiadis et al. [-@hatzigeorgiadisSelfTalkSportsPerformance2011] would have had added little evidence to change prior beliefs, the doubling of research conducted since 2011 also added little evidence to update the prior estimate, and it is likely that much of the research post-2011 (and likely prior to 2011) has been subject to questionable research practices making it unclear whether self-talk interventions affect sports/motor performance or not. This example can act as a warning to researchers in the sport and exercise sciences about what kinds of questions they should ask themselves prior to beginning their work. Such methods as those demonstrated here, when used prospectively, can aid researchers in determining whether further research of a particular experimental intervention is in fact warranted; and when used in retrospect may reveal where research has been a waste. Considering the limited resources and time for conducting research their application may suggest it to be more worthwhile moving onto other more pertinent research questions or programmes. In the case of self-talk specifically, future research might be best placed to continue with the intention of building strong theories regarding the conceptualisation and operationalisation of self-talk and, following this, test if there really is an intervention effect as hypothesised whilst incorporating safe-guards for questionable research practices e.g., pre-registration/registered-reports [@chambersPresentFutureRegistered2022].
